### 2017-02-16 doc/manuscript.Rmd
Made edits based on Tony's comments (discussion only)

We found that in an at-risk for diabetes Canadian population who were mainly female 
and of European ancestry, the specific TAGFA composition was strongly associated 
with insulin sensitivity and moderately associated with beta-cell function. In 
particular, TAGFA myristic acid (14:0), 7-tetradecenoic acid {{ confirm name }} 
(14:1n-7), palmitic acid (16:0), and palmitoleic acid (16:1n-7) all strongly and
negatively predicted lower insulin sensitivity. While most TAGFA were not
associated with beta-cell function, two fatty acids, palmitic acid (16:0) and *cis*-vaccenic acid (18:1n-7), were
associated negatively and positively, respectively, with both measures of 
beta-cell function. We also found that four TAGFA (14:0, 14:1n-7, 16:0, 16:1n-7)
clustered together, and this cluster also strongly predicted lower insulin sensitivity. 
These four fatty acids are also involved in the *de novo* lipogenesis (DNL) of refined and simple 
carbohydrates [@Hodson2008a; @Kawano2013a]. Our results, which are in agreement with the current literature,
suggest that higher activity of DNL (potentially through higher intakes of
simple carbohydrates) may increase the risk for diabetes, likely through
worsening insulin sensitivity.

@@

To our knowledge, no longitudinal study to date has examined the role of the composition
of the TAGFA fraction on detailed OGTT-derived metabolic measures using advanced
{{another word? maybe 'appropriate'?}} statistical techniques designed to maximally
utilize the data. Two large prospective studies have been published that had
similar research objectives as the present analysis. One was a nested
case-control analysis (n=189 cases and n=189 controls) within the Framingham offspring
cohort [@Rhee2011a], which found that TAGFA composition characterized by a lower carbon chain
and fewer double bonds (e.g. 14:0, 16:0) had a higher risk for diabetes after 
12-years while higher carbon chain and more double bond TAGFA had a lower risk 
for diabetes. This pattern of TAGFA with diabetes risk remained the same when 
they examined HOMA-IR cross-sectionally at the baseline visit. The other study 
was a prospective cohort of males in Finland [@Lankinen2015a], for which 
TAGFA data were available for 831 participants after 6-years of follow-up. This 
cohort has similar outcomes as PROMISE, however, OGTT data was only available at
the 6-year visit. As with our results, they found a larger effect size for
insulin sensitivity compared to the beta-cell function. There were some
similarities in results between the present analysis and the Finnish study,
specifically for the saturated fatty acids, docosapentaenoic acid (22:5n-3),
eicosapentaenoic acid (20:5n-3), and arachidonic acid (20:4n-6), though they did
not find any associations with 7-tetradecenoic acid (14:1n-7) and palmitoleic
acid (16:1n-7). Our study extends these findings by using multiple measurements
of metabolic function and using statistical techniques that adequately make use
of the longitudinal and high-dimensional dataset. In another study of a much
smaller (n=16) mostly female group [@Kotronen2009a], the authors report a
positive correlation between total esterified (of which TAG make up the
majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR, similar as our results except
for the 18:1n-9 finding.

Previous research has shown that carbohydrate intake increases DNL 
[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a]. In
particular, DNL from refined or simple carbohydrate sources increases the 14 to
16 chain fatty acids as well as the 18 chain TAGFA. Several studies have shown a
link between higher estimated DNL and an increased risk for metabolic 
dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]. Our study
extends these findings by showing that TAGFA with 14 to 16 carbons clustered together and all
strongly predicted lower insulin sensitivity. While these fatty acids also had a
significant association with beta-cell function, the results were not as strong
as with the insulin sensitivity. However, while the DNL cluster had a strong
harmful association with insulin sensitivity when we include it in the GEE
models, the magnitude of the association was not as large as when using
clinically measured TAG, reinforcing that the total TAG concentration still more
strongly predicts metabolic dysfunction.

The link between higher DNL and increases in specific fatty acids has been
examined in several studies. Previous studies that have examined DNL have used
markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or 16:1n-7
to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are major
limitations to using these ratios as the fatty acids used in their calculation
can also be obtained from the diet in addition to being created through DNL
[@Hodson2008a]. A small, short feeding trial (n=24) was conducted to identify
fatty acids that most accurately reflected DNL, as potential biomarkers
[@Lee2015a]. The study found that palmitoleic acid (16:1n-7), directly measured
DNL using isotopes, and liver fat were all highly correlated with each other
(r>0.50), suggesting that 16:1n-7 may be a good biomarker for hepatic DNL. In
another small (n=14) feeding trial, meal type (high fat vs low fat) was tested
to determine its effect on DNL and TAGFA composition [@Wilke2009a]. The authors
reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
carbohydrate) group. These fatty acids are similar to the fatty acids we found
that clustered together using the PLS analysis, implicating these fatty acids as
indicative of a higher carbohydrate diet. A higher carbohydrate diet, 
particularly one characterized by a predominance of simple carbohydrates, may
lead to greater DNL in an attempt to control blood glucose, thus increasing hepatic
fat stores and consequently increasing the amount of TAGFA in circulation
[@Chehade2013a; @Verges2015a]. The higher concentration of circulating 14 and 16
carbon fatty acids may then expose tissues to greater lipotoxicity, for instance
from palmitic acid (16:0), which is well-known to have harmful effects on
tissues [@Riserus2008a; @Iggman2010a].

We found no interaction by time for any of the TAGFA on the metabolic outcomes.
Analyzing these biological processes longitudinally is challenging due to the
potential feedback mechanisms involved. For instance, while greater DNL may
promote muscle insulin resistance, the reverse may also be true
[@Flannery2012a]. Higher insulin resistance may encourage greater DNL to handle
the higher blood glucose. To illustrate this, in a weight loss intervention trial
(n=19 with TAGFA data), participants who lost weight over 33 weeks showed higher
insulin sensitivity and lower TAGFA composition indicative of lower DNL (less
14:0, 14:1, 16:0, 16:1, etc) [@Schwab2008a]. Given the complex biological
mechanisms and feedback loops involved, disentangling whether insulin
sensitivity influences TAGFA more strongly than TAGFA influencing insulin
sensitivity will require more complicated research designs and analyses. Our own
findings of no interaction by time with the TAGFA tentatively suggest that the
influence of either TAGFA or insulin resistance within the feedback loop may be
of relatively equal strength.

Serum TAG are packaged into lipoproteins, particularly VLDL, by the liver. The
sources of fatty acids for TAG production during fasting are stored fatty acids
within the liver, circulating NEFA, and through DNL [@Barrows2006a]. Therefore, the
TAGFA composition reflects not just dietary intake but also stored fatty acids
from the adipose tissue and the DNL capacity of the individual. During fasting, a
large portion of fatty acids bound together as TAG are obtained from circulating
NEFA. A study examining this found that NEFA contributed 60-80% of newly synthesized
VLDL-TAG during fasting and ~40% during the fed state while dietary fatty acids
contributed ~25% to newly synthesized VLDL-TAG [@Barrows2006a]. In an analysis of the
serum NEFA fraction in the PROMISE cohort with the OGTT-derived outcomes
(under review), we found that higher total NEFA but not the specific
composition associated with lower beta-cell function and not with insulin 
resistance. While NEFA may contribute substantially to the TAG fraction, there 
may be something unique about the specific composition of the TAGFA fraction 
that may allow it to enter insulin sensitive tissues more easily than NEFA. It 
may be that VLDL receptors allow fatty acids to enter these tissue more easily
than NEFA bound to albumin or it may be due to the other influences on TAGFA 
composition such as dietary intake since circulating NEFA primarily come from
adipose tissue [@Verges2015a].

@@

There are a few important limitations to our study. In addition to the
observational nature of the cohort, the biology underlying the interaction
between TAGFA, insulin secretion, and insulin sensitivity is complex and
multifaceted and as such there may be some residual confounding we were
unaware of or had not been able to adjust for. However, we have taken extensive,
empirically based precautions in identifying potential confounders and mediators
through the use of the DAG modeling, relying on previous literature, and through
QIC model fit comparison methods.

@@

PLS is a well-established technique for constructing predictive models of high
dimensionality data structures (i.e. fatty acid composition), however a 
limitation is that the initial models analyzed through PLS and the final 
computed scores are not able to control for potential confounders and other 
effect modifiers. Likewise, PLS is not able to handle longitudinal data so only 
the baseline visit was used in the PLS analysis, though we analyze the extracted
scores using the GEE modeling to overcome this limitation. In light of these
limitations, we still observe concordant results between the PLS and GEE analyses.

Even with these limitations, our study has several notable strengths, including
the longitudinal design and the rigorous statistical techniques and methods
applied in analyzing the data. These statistical techniques take advantage of
the longitudinal data to allow appropriate investigation of temporal
relationships and are able to handle the multivariate nature of the data.
Lastly, our cohort contains highly detailed and comprehensive variable
measurements at each collection visit, as well as having both concentration and
mol% data for the fatty acids.

@@

In conclusion, we found that a TAGFA composition indicative of higher DNL
(containing higher 14:0, 14:1n-7, 16:0, and 16:1n-7) associated strongly with lower
insulin sensitivity and moderately with lower beta-cell function. The fatty
acids that clustered together represent fatty acids created from DNL, which is
characteristic of higher simple carbohydrate (e.g. added sugar) intake. Our
results, which are congruent with current evidence, suggest that higher DNL,
likely due to greater intake of simple or refined carbohydrates, may increase the
risk of diabetes through worsening of insulin sensitivity.

>>>>

In the present study, we found that in a Canadian cohort at risk for diabetes,
several specific TAGFA and groups of TAGFA were strongly associated
with insulin sensitivity and moderately associated with beta-cell function. In 
particular, the TAGFA myristic acid (14:0), 7-tetradecenoic acid {{ confirm name }} 
(14:1n-7), palmitic acid (16:0), and palmitoleic acid (16:1n-7) all strongly and
negatively predicted lower insulin sensitivity. While most TAGFA were not
associated with beta-cell function, two fatty acids, palmitic acid (16:0) and
*cis*-vaccenic acid (18:1n-7), were associated negatively and positively,
respectively, with measures of beta-cell function. Using PLS, we also found that four
TAGFA (14:0, 14:1n-7, 16:0, 16:1n-7) clustered together, and that this cluster
strongly predicted lower insulin sensitivity. These four fatty acids are
involved in the *de novo* lipogenesis (DNL) of refined and simple carbohydrates
[@Hodson2008a; @Kawano2013a]. Our results suggest that higher activity of DNL
(potentially through higher intakes of simple carbohydrates) may increase the
risk for diabetes, primarily through worsening insulin sensitivity.

@@

To our knowledge, no longitudinal study to date has examined the role of the
composition of the TAGFA fraction on detailed OGTT-derived metabolic measures 
Two large prospective studies have been published that similarly examined TAGFA 
composition and diabetes outcomes. Rhee *et al* presented a nested case-control
analysis (n=189 cases and n=189 controls) within the Framingham offspring cohort
[@Rhee2011a], which found that subjects with a TAGFA composition characterized
by a lower carbon chain and fewer double bonds (e.g. 14:0, 16:0) had a higher
risk for diabetes after 12-years while those with a profile characterized by
higher carbon chain and more double bond TAGFA had a lower risk for diabetes.
A similar pattern of TAGFA was also associationed with HOMA-IR cross-sectionally
at the baseline visit. In addition, Lankinen *et al* reported on
a prospective cohort of males in Finland [@Lankinen2015a], for which 
TAGFA data were available for 831 participants after 6-years of follow-up. 
They found that most saturated fatty acids had negative , docosapentaenoic acid (22:5n-3),
eicosapentaenoic acid (20:5n-3), and arachidonic acid (20:4n-6), though they did
not find any associations with 7-tetradecenoic acid (14:1n-7) and palmitoleic
acid (16:1n-7). {{ fix up }}
There were some
similarities in results between the present analysis and the Finnish study,
Our study extends these findings by using multiple measurements
of metabolic function and as well as multivariate statistical approaches that
allowed us to identify a cluster of TAGFA.
In another study of a much
smaller (n=16) mostly female group [@Kotronen2009a], the authors report a
positive correlation between total esterified (of which TAG make up the
majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR, similar as our results except
for the 18:1n-9 finding.

Previous research has shown that carbohydrate intake increases DNL 
[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a]. In
particular, DNL from refined or simple carbohydrate sources increases the 14 to
16 chain fatty acids as well as the 18 chain TAGFA. Several studies have shown a
link between higher estimated DNL and an increased risk for metabolic 
dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]. Our study 
extends these findings by showing that TAGFA with 14 to 16 carbons clustered
together and this pattern strongly predicted lower insulin sensitivity. While
these fatty acids also had a significant association with beta-cell function,
the magnitude of associations were more modest compared to those for insulin
sensitivity.

The link between higher DNL and increases in specific fatty acids has been
examined in several studies. Previous studies that have examined DNL have used
markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or 16:1n-7
to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are 
limitations to using these ratios as the fatty acids used in their calculation
can also be obtained from the diet in addition to being created through DNL
[@Hodson2008a]. A feeding trial (n=24) was conducted to identify 
the fatty acids that most accurately reflected DNL as potential biomarkers
[@Lee2015a]. The study found that palmitoleic acid (16:1n-7), directly measured
DNL using isotopes, and liver fat were all highly correlated with each other
(r>0.50), suggesting that 16:1n-7 may be a good biomarker for hepatic DNL. In
another small (n=14) feeding trial, meal type (high fat vs low fat) was tested
to determine its effect on DNL and TAGFA composition [@Wilke2009a]. The authors
reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
carbohydrate) group. These fatty acids are similar to the fatty acids we found
that clustered together using the PLS analysis, implicating these fatty acids as
indicative of a higher carbohydrate diet. A higher carbohydrate diet, 
particularly one characterized by a predominance of simple carbohydrates, may
lead to greater DNL in an attempt to control blood glucose, thus increasing hepatic
fat stores and consequently increasing the amount of TAGFA in circulation
[@Chehade2013a; @Verges2015a]. The higher concentration of circulating 14 and 16
carbon fatty acids may then expose tissues to greater lipotoxicity, for instance
from palmitic acid (16:0), which is well-known to have harmful effects on
tissues [@Riserus2008a; @Iggman2010a].

{{fix this}}
However, while this DNL cluster had a strong negative association
with insulin sensitivity, the magnitude of the association was slightly weaker
compared to clinically measured TAG.

The direction of association between TAGFA and insulin sensitivity is unclear 
from cross-sectional studies due to the physiological feedback mechanisms 
involved. For example, while greater DNL may promote muscle insulin resistance, 
the reverse may also be true [@Flannery2012a]. Higher insulin resistance may
encourage greater DNL to handle the higher blood glucose. To illustrate this, in
a weight loss intervention trial (n=19 with TAGFA data), participants who lost
weight over 33 weeks showed higher insulin sensitivity and lower TAGFA
composition indicative of lower DNL (less 14:0, 14:1, 16:0, 16:1, etc)
[@Schwab2008a]. Given the complex biological mechanisms and feedback loops
involved, disentangling whether insulin sensitivity influences TAGFA more
strongly than TAGFA influencing insulin sensitivity will require more
complicated research designs and analyses. While further research will need to
confirm this, the lack of a time interaction we observed in addition to the
timing of our measures likely suggests that it is the TAGFA predicting insulin
resistance.

@@

Our study has potential limitations that need to be considered when interpreting
the results. Firstly, this is an observational cohort and as such there may be
some residual confounding we were not able to control for or were unaware of. 
However, we have taken extensive, empirically based precautions in identifying
potential confounders and mediators through the use of the DAG modeling, relying
on previous literature, and through QIC model fit comparison methods.

@@

PLS is a well-established technique for constructing predictive models of high
dimensionality data structures (i.e. fatty acid composition), however a 
limitation is that the initial models analyzed through PLS and the final 
computed scores are not able to control for potential confounders and other 
effect modifiers. Likewise, PLS is not able to handle longitudinal data so only 
the baseline visit was used in the PLS analysis, although we analyzed the extracted
scores using the GEE modeling to overcome this limitation and observed
concordant results between the PLS and GEE analyses.

Our study has several notable strengths, including the longitudinal design and
the use of advanced statistical techniques for data analysis. These statistical
techniques take advantage of the longitudinal data to allow appropriate
investigation of temporal relationships and are able to handle the
multidimensional nature of the data. Lastly, our cohort contains highly detailed
and comprehensive variable measurements for the fatty acids and outcomes,
of which were collected at each visit.

@@

In conclusion, we found that a TAGFA composition indicative of higher DNL
(containing higher 14:0, 14:1n-7, 16:0, and 16:1n-7) associated strongly with lower
insulin sensitivity and (more moderately) with lower beta-cell function. The fatty
acids that clustered together represent fatty acids created from DNL, which is
characteristic of higher simple carbohydrate (e.g. added sugar) intake. Our
results, which are congruent with current evidence, suggest that higher DNL,
likely due to greater intake of simple or refined carbohydrates, may increase the
risk of diabetes through worsening of insulin sensitivity.
